---
title: "exercise-2"
author: "Matthew McAvoy"
date: "August 25, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load XML file

Requires XML to read in xml files.
```{r}
require("XML")
```

Sets the working directory and loads the xml file into local memory.
```{r}
setwd("C:/Users/homur/OneDrive/New College/EDA/Week 1")
courseData <- xmlParse("reed-courses.xml")
```

xmlToList to convert the xml file into a more usable form, topxml access's the top node, then xmlSApply to extract xml values.
```{r}
courseXml <- xmlToList(courseData)
topxml <- xmlRoot(courseData)
topxml <- xmlSApply(topxml, function(x) xmlSApply(x,xmlValue))
```

If we were to inspect it now, it would look like an xml file; meaning not very pretty. The next step converts the xml file into an R data frame for further operations.
```{r}
xml_df <- data.frame(t(topxml))
head(xml_df)
```

Now upon inspection, it looks much better as a data frame. We then use the str() function to look at metadata of the xml file.
```{r}
str(xml_df)
```

From this, we can answer some of the questions.

2. There are 31 distinct subjects listed in the document.

4. There are 136 distinct insturctors listed.

Using the following summary, we see an empty field in instructor, which allows us to answer number 3.

3. empty row corresponds to professor names of NULL = 15.

```{r}
summary(xml_df)
```









