---
title: "Getting started with R"
author: "P. McDonald"
date: "August 25, 2016"
transition: rotate
incremental: true
output: ioslides_presentation
---

##Last time

*    An introduction to terms
*    Course description
*    R Markdown
*    Reading files in R
    +   Example: reading csv
        
##Solution to first exercise, part 1

Here is a script which downloads the required csv and reads it into memory:

```{r, eval=FALSE}
## Pat McDonald   08/23/2016
## Script to load a csv file to memory

##assign url
url <-  "https://data.somervillema.gov/api/views/j2bq-38ev/rows.csv"

##read the file

TrafficData.csv <- read.csv(url, header=TRUE)

```

##Doing a little more

We can download the data and read it into memory
```{r,eval=FALSE}
##continue previous code to download a csv file

##download to appropriate directory
download.file(url, "~/Classes/16-17/Fall/EDA2016/Data/SMATrafficViolation.csv")

##read file
TrafficData.csv <- read.csv("~/Classes/16-17/Fall/EDA2016/Data/SMATrafficViolation.csv", header=TRUE)

##Provenance
SMATrafficViolationProvenance <- paste("Downloaded from", url, "At", Sys.time(), sep = " ")
write(SMATrafficViolationProvenance, file="~/Classes/16-17/Fall/EDA2016/Data/SMATrafficViolation")
```

##Solution to first exercise, part 2

This problem was a little more challenging.

##Solution to first exercise, part 2

This problem was a little more challenging.

Assuming you did the first part, there is an obvious structural change which must occur:  it is unlikely that read.csv will work for json files.

There are two ways to proceed:

##Solution to first exercise, part 2

This problem was a little more challenging.

Assuming you did the first part, there is an obvious structural change which must occur:  it is unlikely that read.csv will work for json files.

There are two ways to proceed:

1.    Read the R manual

##Solution to first exercise, part 2

This problem was a little more challenging.

Assuming you did the first part, there is an obvious structural change which must occur:  it is unlikely that read.csv will work for json files.

There are two ways to proceed:

1.    Read the R manual
2.    Google something like "read JSON in R"

##Solution to first exercise, part 2

This problem was a little more challenging.

Assuming you did the first part, there is an obvious structural change which must occur:  it is unlikely that read.csv will work for json files.

There are two ways to proceed:

1.    Read the R manual
2.    Google something like "read JSON in R"

The second option is easier.  

##Google to the rescue

Run the google search and look for a stack exchange answer.

Under the title "Parse JSON with R", you will find a number of replies.

Start with the one having the most up-votes.

##Google to the rescue

Run the google search and look for a stack exchange answer.

Under the title "Parse JSON with R", you will find a number of replies.

Start with the one having the most up-votes.

What it says is:

1.    There is a package that will solve the problem
2.    There is an update which suggests that the second package is even more straightforward.

##Google to the rescue

Run the google search and look for a stack exchange answer.

Under the title "Parse JSON with R", you will find a number of replies.

Start with the one having the most up-votes.

What it says is:

1.    There is a package that will solve the problem
2.    There is an update which suggests that the second package is even more straightforward.

The second most up-votes also references the second package.

##Installing packages

Installing packages in R is really easy - especially from R Studio.  

Just select packages in the lower right corner, select install, type the name of the package (enable dependencies) and your done.

It's also easy to install from the R command line (google it).

##Code for exercise 1, part 2

```{r, eval = FALSE}
## Pat McDonald   08/23/2016
## Script to download a json file and load to memory

##load required library
library(jsonlite)
##assign url
url <-  "https://data.somervillema.gov/api/views/j2bq-38ev/rows.json"

##read file
TrafficData.json <- fromJSON(txt=url)

save(TrafficData.json, file="~/Classes/16-17/Fall/EDA2016/Data/SMATrafficViolation.json")

##Provenance
SMATrafficViolationJSONProvenance <- paste("Downloaded from", url, "At", Sys.time(), sep = " ")
write(SMATrafficViolationJSONProvenance, file="~/Classes/16-17/Fall/EDA2016/Data/SMATrafficViolationJSON")
```

##Other tools for reading data

There are a variety of tools for importing data into R.

read.table is a function which subsumes a number of other functions.

##Other tools for reading data

There are a variety of tools for importing data into R.

read.table is a function which subsumes a number of other functions.

For example, it can be used to download csv files:

```{r, eval=FALSE}
url <-  "https://data.somervillema.gov/api/views/j2bq-38ev/rows.csv"

##read the file

TrafficData.csv <- read.table(url, header=TRUE, sep=",")
```
You can find additional information at the man pages and in a variety of tutorials, including 

https://www.datacamp.com/community/tutorials/r-data-import-tutorial#gs.aKMomYE 

##What to do when the data is big

One of the unfortunate limitations of R is its handling of large data.

Reading to memory is problematic for a variety of reasons.

##What to do when the data is big

One of the unfortunate limitations of R is its handling of large data.

Reading to memory is problematic for a variety of reasons.

There are a number of new packages which address the well known limitations (eg, data.table)

##What to do when the data is big

One of the unfortunate limitations of R is its handling of large data.

Reading to memory is problematic for a variety of reasons.

There are a number of new packages which address the well known limitations (eg, data.table)


A nice source of information for importing large data sets into R can be found on the Rblogger site:

https://www.r-bloggers.com/importing-data-into-r-part-two/

##RTM

In th old days, there were more letters:

https://cran.r-project.org/doc/manuals/r-release/R-data.pdf



##Next step: understanding what we have in memory

Every language, R included, has an associated collection of atomic data types.

##Next step: understanding what we have in memory

Every language, R included, has an associated collection of atomic data types.

For R, these types are: numeric, integer, character, logical and factor.

The data structures of R are built on top of its atomic data types.

##Next step: understanding what we have in memory

Every language, R included, has an associated collection of atomic data types.

For R, these types are: numeric, integer, character, logical and factor (actually an attribute setting).

The data structures of R are built on top of its atomic data types.

There are four primary data structures that we will use during the course of our study: vectors, matrices, data frames and lists.

##Next step: understanding what we have in memory

Every language, R included, has an associated collection of atomic data types.

For R, these types are: numeric, integer, character, logical and factor (actually an attribute setting).

The data structures of R are built on top of its atomic data types.

There are four primary data structures that we will use during the course of our study: vectors, matrices, data frames and lists.

Here is a good reference:

http://adv-r.had.co.nz/Data-structures.html



##Keeping track of data structures

It's very important to keep these data structures straight.  

There are a couple of useful functions for doing this:

##Keeping track of data structures

It's very important to keep these data structures straight.  

There are a couple of useful functions for doing this:

*    The class function, *class()*, which returns the class of the R object which is its argument and 
*    The structure function, *str()*, which provides a much richer description of its argument.  
*    The summary function, *summary()*, which provides various statistical model fit summaries of R objects.

https://www.r-bloggers.com/exploratory-data-analysis-useful-r-functions-for-exploring-a-data-frame/

##Data structures: vectors

The most important data structure of R is a *vector.*  A vector is a tuple whose elements consist of a single atomic data type.

```{r}
v1 <- c("h")
v2 <- c("h","t")
v3 <- c("ht")
```

The str() function gives us information on the structure of the above

##Using the structure function

```{r}
str(v1) 
str(v2)
str(v3)
```

Thus, each of the above is a vector with v1 and v3 being atomic.


##Vectors and type conversion

Thus, its easy to guess the structure of one of the objects currently in our working memory:

```{r, eval=FALSE}
str(url)
```

##Vectors and type conversion

Thus, its easy to guess the structure of one of the objects currently in our working memory:

```{r, eval=FALSE}
str(url)
```

Suppose we change a component of v2:

```{r}
v2[2] <- 1
v2
```

##Vectors and type conversion

Thus, its easy to guess the structure of one of the objects currently in our working memory:

```{r, eval=FALSE}
str(url)
```

Suppose we change a component of v2:

```{r}
v2[2] <- 1
v2
```

Note what has happened: R has interpreted our change to mean that we intended to change the character "t" to the character "1".  

##Vectors and type conversion

Thus, its easy to guess the structure of one of the objects currently in our working memory:

```{r, eval=FALSE}
str(url)
```

Suppose we change a component of v2:

```{r}
v2[2] <- 1
v2
```

Note what has happened: R has interpreted our change to mean that we intended to change the character "t" to the character "1".  

This is called *coercion* and it can lead to problems.


##Vectors and vectorization

R is *vectorized* language: most of the functions of R are designed to take vectors as an input and act on the components of the input.  

##Vectors and vectorization

R is *vectorized* language: most of the functions of R are designed to take vectors as an input and act on the components of the input.  

R does this very quickly: if you aren't exploiting vectorization then with very high probability, you are not optimizing.

##Matrices

Matrices are collections of vectors which are all of the same type.

##Matrices

Matrices are collections of vectors which are all of the same type.

Here is an example of a matrix built from this year's Calculus Readiness Test results:

```{r}
load("~/Classes/16-17/Fall/EDA2016/Data/M1")
str(M1)
head(M1)
```

##Matrices
We can look at the first row of M1

```{r}
str(M1[1,])
```

It is a vector of integers of length 8.

Similarly, the 3 column is a vector of integers of length 107:

```{r}
str(M1[,3])
```

##Introduction to subsetting

For the above matrix example we specified a particular row or column and got a vector.  

##Introduction to subsetting

For the above matrix example we specified a particular row or column and got a vector.  

We can specify a collection of rows or columns:

```{r}
str(M1[,1:3])
```

##Introduction to subsetting
For the above matrix example we specified a particular row or column and got a vector.  
We can specify a collection of rows or columns:
```{r}
str(M1[,1:3])
```
This is a matrix with 107 rows and 3 columns:
```{r}
dim(M1[,1:3])
str(M1[,1:3])
```

##Guess dimensions

```{r, eval=FALSE}
M1[2:4,1:3]
```

##Guess dimensions

```{r, eval=FALSE}
M1[2:4,1:3]
```

```{r}
str(M1[2:4,1:3])
```

##Guess dimensions

```{r, eval=FALSE}
M1[2:4,1:3]
```

```{r}
str(M1[2:4,1:3])
```

```{r}
M1[2:4,1:3]
```

##Data frames

Here is another example from this year's CRT
```{r}
load("~/Classes/16-17/Fall/EDA2016/Data/DF1")
str(DF1)
```

##Data frames

This is an example of a *data frame*: a collection of vectors of the same length, but possibly different data type.

In the above, we have (column) vectors which are integer and (column) vectors which are factor.  

##Data frames

This is an example of a *data frame*: a collection of vectors of the same length, but possibly different data type.

In the above, we have (column) vectors which are integer and (column) vectors which are factor.  

As with matrices, we can refer to columns by number:

```{r}
str(DF1[,2])
```

##Data frames

This is an example of a *data frame*: a collection of vectors of the same length, but possibly different data type.

In the above, we have (column) vectors which are integer and (column) vectors which are factor.  

As with matrices, we can refer to columns by number:

```{r}
str(DF1[,2])
```

We can also refer to columns by name:

```{r}
str(DF1$Percent)
```

##A note on factor types

Factors are used to represent categorical values.  

##A note on factor types

Factors are used to represent categorical values.  

A factor vector has components which take values in a given set (for example, race or sex or percent correct).  

##A note on factor types

Factors are used to represent categorical values.  

A factor vector has components which take values in a given set (for example, race or sex or percent correct).  

R codes the values with integers to speed analysis.

Example:

```{r}
head(DF1$Percent)
```

##A note on factor types

Example:

```{r}
head(DF1$Percent)
```

We can see what's behind the symbol:

```{r}
head(as.numeric(DF1$Percent))
```

This is a useful introduction:  https://www.r-bloggers.com/data-types-part-3-factors/

##Data frames

We have seen examples of data frames before:

```{r}
SMATrafficViolation.csv <- read.csv("~/Classes/16-17/Fall/EDA2016/Data/SMATrafficViolation.csv")
str(SMATrafficViolation.csv)
```

Thus, SMATrafficViolation.csv is also a *data frame.*

##Lists

A list is a vector of objects:
```{r}
ListEx <- list(list(DF1), DF1,v2)
str(ListEx)
```


##Lists

A list is a vector of objects:
```{r}
ListEx <- list(list(DF1), DF1,v2)
class(ListEx)
```

We have already seen an example:

```{r, eval=FALSE}
class (TrafficData.json)
```

##Lists

We have already seen an example:

```{r, eval=FALSE}
class (TrafficData.json)
```

What about 
```{r, eval=FALSE}
DF1[2,]
```

##Lists

We have already seen an example:

```{r, eval=FALSE}
class (TrafficData.json)
```

```{r}
DF1[2,]
```

##Lists

We have already seen an example:

```{r, eval=FALSE}
class (TrafficData.json)
```

```{r}
DF1[2,]
```

We can coerce this to be a list:

```{r}
str(as.list(DF1[2,]))
```

##Lists

Now we can now give an example of why it is important to keep track of data structures via examples we have already implicitly worked out.  

Suppose we try to write DF1 to memory:

```{r,eval=FALSE}
write(DF1, file = "~/Classes/16-17/Fall/EDA2016/Data/EXP")
```

##Lists

```{r,eval=FALSE}
write(DF1, file = "~/Classes/16-17/Fall/EDA2016/Data/EXP")
```

This will throw an error.  The reason is that the factor variable is intrepreted as a list.

This is why we used "save".

##Lists

```{r,eval=FALSE}
write(DF1, file = "~/Classes/16-17/Fall/EDA2016/Data/EXP")
```

This will throw an error.  The reason is that the factor variable is interpreted as a list.

This is why we used "save".

"save"" is not as fast as "write", but it takes a variety of R objects.

