---
title: "exercise-3"
author: "Matthew McAvoy"
date: "August 30, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 3

#```{r}
#library(XML)
#```


We begin by changing to the working directory and reading in the Diabetes data.
```{r}
setwd("C:/Users/homur/OneDrive/New College/EDA/Week 2/dataset_diabetes/dataset_diabetes/")
DiabetesData <- read.csv(file="diabetic_data.csv",head=TRUE,sep=",")
```

## 2. 
The following commands converts our file into a data frame, then looks for all the empty values in the data.
```{r}
DiabetesDataFrame <- data.frame(DiabetesData)
#str(DiabetesDataFrame)
emptyvals <- sapply(DiabetesDataFrame, is.null)
emptyvals
```

## 3. 
We subset on number-emergencies and sum it. Along with finding the number of entries in the original data, we can divide the two to acquire the percent admitted.
```{r}
subsetAdmissions <- DiabetesDataFrame["number_emergency"]

x1 <- nrow(DiabetesDataFrame) # 101766 rows, each for a patient
x2 <- sum(subsetAdmissions) # 20133 patients that have been admitted
percentAdmissions <- x2/x1
```

```{r}
cat('The percentage of admitted patients from the emergency room is', percentAdmissions)
```


## 4. 
To find the mode, we need to use a function. Found online at 'http://www.tutorialspoint.com/r/r_mean_median_mode.htm'
```{r}
# Create the function.
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
```

Then vectorizing and applying the function.
```{r}
modein <- getmode(DiabetesDataFrame$"admission_source_id")
modeout <- getmode(DiabetesDataFrame$"discharge_disposition_id")
```

```{r}
cat('We see the most common admission-id is', modein, 'and the most common discharge-id is', modeout)
```


Next we want to make a new data frame with admission and discharge ids so that we can select on admission id's. Then we vectorize on discharge id and find it's mode.
```{r}
newFrame <- DiabetesDataFrame[,c("admission_source_id","discharge_disposition_id")]
newSubset <- subset(newFrame, admission_source_id == 7)
mode3 <- getmode(newFrame$"discharge_disposition_id")
```

```{r}
cat('We find the mode for this is', mode3, 'again.')
```

5. To characterize the distribution, we can vectorize on admission source and run boxplot().

```{r}
 boxplot(DiabetesDataFrame$admission_source_id)
```





