---
title: "Getting started with R, II"
author: "P. McDonald"
date: "August 29, 2016"
transition: rotate
incremental: true
output: ioslides_presentation
---

##Last time

*    Data types
     +    character, integer, numeric, logical, factor
*    Data structures
     +    vectors, matrices, data frames, lists
*    Tools to extract structural information
     +    class, str, summary
     +    subsetting
*    Vectorization     
        
##Solution to second exercise

You were asked to extract some information from an XML file.

Step 1 is to figure out how to read the file.  Google helps.  

```{r}
## Pat McDonald   08/23/2016
## Script to download a XML file and load to memory

##load required library
library(XML)

##assign url
url <-  "http://www.cs.washington.edu/research/xmldatasets/data/courses/reed.xml"

##read file
Reed.xml <- xmlTreeParse(url)

```

##Solution to second exercise
Next, determine the structure of the file you have read into memory:

```{r}
##continuing solution of exercise 2
class(Reed.xml)
```
Note that the return is not a description of one of the data structures described in the previous lecture.

##Solution to second exercise
Next, determine the structure of the file you have read into memory:

```{r}
##continuing solution of exercise 2
class(Reed.xml)
```

Note that the return is not a description of one of the data structures described in the previous lecture.

Google something useful: "converting xml to a data frame in R"

##Solution to second exercise
Next, determine the structure of the file you have read into memory:

```{r}
##continuing solution of exercise 2
class(Reed.xml)
```

Note that the return is not a description of one of the data structures described in the previous lecture.

Google something useful: "converting xml to a data frame in R"

The first stackoverflow entry gives a function which might work ("xmlToDataFrame()"), but suggests there may be problems. 

##Solution to second exercise

Stackoverflow suggests that we convert the XML document to a list:

```{r}
##continuing exercise 2
##convert to list
LReed <- xmlToList(Reed.xml)
class(LReed)
```

##Solution to second exercise

Stackoverflow suggests that we convert the XML document to a list:

```{r}
##continuing exercise 2
##convert to list
LReed <- xmlToList(Reed.xml)
class(LReed)
```

We can extract a little more information using the "dim" tool:

```{r}
##continuing exercise 2
dim(LReed)
```

##Solution to second exercise

Thus, the matrix LReed has `r nrow(LReed)` rows and `r ncol(LReed)` columns.  

The name of a row or a column is a useful piece of information.  Odds are that there is a tool to extract such information.  Google it.

##Solution to second exercise

Thus, the matrix LReed has `r nrow(LReed)` rows and `r ncol(LReed)` columns.  

The name of a row or a column is a useful piece of information.  Odds are that there is a tool to extract such information.  Google it.

```{r}
##continuing exercise 2
rownames(LReed)
```

##Solution to second exercise
Now we can subset to extract the required information.

Begin by extracting (subsetting) the row corresponding to subjects.  There are a couple of ways to do this.

```{r}
##continuing exercise 2
ReedSubjects <- LReed[2,]
class(ReedSubjects)
str(ReedSubjects)
```

##Solution to second exercise

It looks as though we are on the correct path.  What we need is a tool which "extracts unique entries" from our list.

Again, this would be something useful to lots of people, so there is probably already a utility that solves the problem.

##Solution to second exercise

It looks as though we are on the correct path.  What we need is a tool which "extracts unique entries" from our list.

Again, this would be something useful to lots of people, so there is probably already a utility that solves the problem.

Google "extracting unique entries in R".  

##Solution to second exercise

It looks as though we are on the correct path.  What we need is a tool which "extracts unique entries" from our list.

Again, this would be something useful to lots of people, so there is probably already a utility that solves the problem.

Google "extracting unique entries in R".  

https://stat.ethz.ch/R-manual/R-devel/library/base/html/unique.html

##Solution to second exercise

It looks as though we are on the correct path.  What we need is a tool which "extracts unique entries" from our list.

Again, this would be something useful to lots of people, so there is probably already a utility that solves the problem.

Google "extracting unique entries in R".  

https://stat.ethz.ch/R-manual/R-devel/library/base/html/unique.html

The function "unique" takes a vector or data frame as an input.  

We have a list.

There is a tool to solve this problem:

##Solution to second exercise

The command "unlist" converts a list to a vector consisting of the atomic components of the list.

```{r}
##continuing exercise 2
RS <- unlist(ReedSubjects)
str(RS)
```

##Solution to second exercise

The command "unlist" converts a list to a vector consisting of the atomic components of the list.

```{r}
##continuing exercise 2
RS <- unlist(ReedSubjects)
str(RS)
```

Now we just compute the number of unique subjects by computing the length of the vector of unique entries in "RS"

```{r}
##continuing exercise 2
length(unique(RS))
```

##Solution to second exercise

Our approach to the first problem suggests we try a similar approach to the second problem:

```{r}
##continuing exercise 2
##emulate
ReedInstructors <- LReed[7,]
class(ReedInstructors)
```

##Solution to second exercise

We can take a look at the first few entries:
```{r}
##continuing exercise 2
head(ReedInstructors)
```

##Solution to second exercise

Proceeding as before:

```{r}
##continuing exercise 2
RI <- unlist(ReedInstructors)
str(RI)
```

This is a little odd:  There are fifteen fewer components in the instructor vector than there were in the subject vector.

```{r}
head(RI)
```

##Solution to the second exercise

It seems that there is an unintended consequence of using unlist:  It truncates NULL entries.  

##Solution to the second exercise

It seems that there is an unintended consequence of using unlist:  It truncates NULL entries.  

Since we started with `r length(ReedInstructors)` columns and ended with `r length(RI)` columns, we know that there are precisely `r length(ReedInstructors)-length(RI)` NULL entries.

##Solution to the second exercise

It seems that there is an unintended consequence of using unlist:  It truncates NULL entries.  

Since we started with `r length(ReedInstructors)` columns and ended with `r length(RI)` columns, we know that there are precisely `r length(ReedInstructors)-length(RI)` NULL entries.

Similarly, we can compute the number of unique instructor entries (not counting NULL as an entry)

```{r}
length(unique(RI))
```

##A little munging

There are often reasons to keep a marker for the place where a NULL value occurs.  To do so, google the phenomena:

##A little munging

There are often reasons to keep a marker for the place where a NULL value occurs.  To do so, google the phenomena:

There are a couple of options for inserting NA for NULL: one approach is to use coercion while the other is to use a feature of vectorization.

##A little munging

There are often reasons to keep a marker for the place where a NULL value occurs.  To do so, google the phenomena:

There are a couple of options for inserting NA for NULL: one approach is to use coercion while the other is to use a feature of vectorization.

We will go with the latter.

##A little munging

There are often reasons to keep a marker for the place where a NULL value occurs.  To do so, google the phenomena:

There are a couple of options for inserting NA for NULL: one approach is to use coercion while the other is to use a feature of vectorization.

We will go with the latter.

To do so requires us to introduce some new tricks.

##Functions

There are a number of built-in functions which can be very useful:

http://www.statmethods.net/management/functions.html

##Functions

There are a number of built-in functions which can be very useful:

http://www.statmethods.net/management/functions.html

It is also possible to define your own functions:

http://www.statmethods.net/management/userfunctions.html

##Functions

There are a number of built-in functions which can be very useful:

http://www.statmethods.net/management/functions.html

It is also possible to define your own functions:

http://www.statmethods.net/management/userfunctions.html

In addition to these options, there is a suite of functions which allow a user to perform a given function to every element of a given data structure.  

##Functions

There are a number of built-in functions which can be very useful:

http://www.statmethods.net/management/functions.html

It is also possible to define your own functions:

http://www.statmethods.net/management/userfunctions.html

In addition to these options, there is a suite of functions which allow a user to perform a given function to every element of a given data structure.  

This suite is important: it allows users to exploit vectorization.

##sapply
The function sapply provides an example of how to leverage vectorization.

##sapply
The function sapply provides an example of how to leverage vectorization.

sapply examines every entry of a list, performs a prescribed action on the element, and returns a vector of results.

##sapply
The function sapply provides an example of how to leverage vectorization.

sapply examines every entry of a list, performs a prescribed action on the element, and returns a vector of results.

Consider the following code:

```{r}
##determine the location of NULL entries, create a 
##logical vector which encodes the data

labels <- sapply(ReedInstructors,is.null)
class(labels)
```

##Using labels

We can use the labels to mark locations where we will insert NA to replace NULL.

```{r}
##copy NA over locations where NULL is located and unlist

RIc <- ReedInstructors
RIc[labels] <- NA
RIwNA <- unlist(RIc)
str(RIwNA)
```

##A second solution to second exercise

Now we can finish the problem

We already know the number of NULL instructors:
```{r}
sum(labels)
```
Thus, there are `r sum(labels)` courses with NULL as the listed instructor.

##A second solution to second exercise

Similarly,
```{r}
length(unique(RIwNA))
length(unique(RI))
```
and there are `r length(unique(RI))` unique instructors where we disregard NULL or NA entries.

##The apply suite

The apply suite consists of three functions: apply, sapply, and lapply.

##The apply suite

The apply suite consists of three functions: apply, sapply, and lapply.

There are many interesting examples (and some not so interesting) of how one might use the apply suite to get things done.

##The apply suite

The apply suite consists of three functions: apply, sapply, and lapply.

There are many interesting examples (and some not so interesting) of how one might use the apply suite to get things done.

Which element of the suite you use depends on how the input is structured, what the output is to look like, and what is to be done.

For details, here is a good place to start:

https://www.r-bloggers.com/using-apply-sapply-lapply-in-r/

##Flow control

It is possible to avoid using the apply suite, choosing instead to use conditionals and looping:

```{r}
##loops to create labels

L <- as.logical(rep(0,703))
for (i in 1:length(ReedInstructors)) { 
  if (is.null(ReedInstructors[[i]]) ) {
    L[i] <- TRUE}
}
sum(L)
```

##Flow control

It is possible to avoid using the apply suite, choosing instead to use conditionals and looping:

```{r}
##loops to create labels

L <- as.logical(rep(0,length(ReedInstructors)))
for (i in 1:length(ReedInstructors)) { 
  if (is.null(ReedInstructors[[i]]) ) {
    L[i] <- TRUE}
}
sum(L)
```

This is not a good idea.

For more on syntax for flow control:  http://www.statmethods.net/management/controlstructures.html

##Another problem

For what collection of subjects is there an entry of NULL for at least one  instructor? 

We have the required data, we just need to format things in a fashion which makes answering the question easy.

One option is to just create a data frame and subset:

```{r}
##build a data frame
DFR <- data.frame(RS,RIwNA)
str(DFR)
```

##Another problem

Subset to extract the appropriate vector:

```{r}
##subsetting to extract the correct data
NullSubjects <- DFR[is.na(RIwNA),1]
##summarize the outcome
summary(NullSubjects)
```
##Another problem

Finally, answer the question

```{r}
##print the result to screen
cat('There are', length(unique(NullSubjects)), 'distinct subjects that 
    have at least one instance of a NULL instructor')
Answer <- paste('There are', length(unique(NullSubjects)), 'distinct subjects that have at least one instance of a NULL instructor', sep=" ")
```

```{r}
Answer
```

##Yet more fun
Being able to count is important.

##Yet more fun
Being able to count is important.

How many distinct math instructors offered courses?

##Yet more fun
Being able to count is important.

How many distinct math instructors offered courses?

```{r}
##to count: first subset, then unique, then length
length(unique(DFR[which(RS == "MATH"),2]))
```



